 

### 2021-07-26 vertical-thermal-review.Rmd
including edits from google doc

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
Foliage distribution is structured in correspondence with the distribution of tree sizes within a forest, where leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs. `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`). 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In open canopy forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig. `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations followed by disturbance all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019]. 
In this review, we focus on growing season conditions (peak LAD) unless otherwise noted.

**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the profile shape modified by leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Figs. `r fig_NEON_vertical`b-d, `r SIfig_NEON_all`, @koike_leaf_2001].** 
This gradient is more pronounced under dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than under more open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs. `r fig_NEON_vertical`d, `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].

**Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018] to open forests and savannas [@curtis_intracanopy_2019; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs. `r fig_NEON_vertical`e, `r SIfig_NEON_all`). 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to 4.1 m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds averaged $\le$ 1.1 m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).

**Atmospheric concentrations of carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.**
Atmospheric CO~2~ concentrations tend to be higher near ground level at night, associated with plant and soil respiration, but the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999]. Elevated CO~2~ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are negligible for most of day when photosynthesis is occurring, pattern in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.

Relative humidity, RH, tends to be higher in the understory and decreases with height across the vertical profile, although this trend is absent in open forests [Figs. `r fig_NEON_vertical`f, `r SIfig_NEON_all`, e.g., @mcgregor_tree_nodate].
Dense-canopy forests maintain higher daily maximum RH below-canopy than open forests and nearby open areas, and this effect is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a]. 
In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. Maximum $T_{air}$ and VPD in the understory are further decreased with increasing canopy height and LAI [@juckerCanopyStructureTopography2018]. *( This last sentence needs work.)*

@@

**The buffering patterns described above vary geographically, seasonally, and across forests types [Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`, @frenneForestMicroclimatesClimate2021].**
The strength of buffering increases with canopy density -- for example, across gradients of elevation [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @juckerCanopyStructureTopography2018], moisture, and proximity to water, for example in riparian forests [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019].
Seasonally, canopy buffering capacity is greater in wet than dry seasons, presumably because leaf area is greater in wet seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a].

@@

Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections) with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020]. 
Leaves can be either cooler than air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ [megathermy, @doughtyAreTropicalForests2008].
While $T_{leaf}$ is most commonly within a few degrees of $T_{air}$, it is rarely exactly equal [poikilothermy, @cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

@@

**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based via biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input from radiation minus heat lost to the environment [Fig. `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].**
The dominant component of a leaf's energy balance is net radiation ($R_n$, including shortwave and longwave). High radiation loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig. `r fig_leaf_T`a).
The dominant avenue of heat loss to the environment is latent heat flux ($\lambda E$), wherein the loss of energy required to convert water from a liquid to a vapour state during evaporation and transpiration cools the leaf. 
Evapotranspiration ($E$) is controlled primarily by stomatal and boundary layer conductances, $g_s$ and $g_a$, and by the water vapour concentration difference between the leaf surface and the bulk air.
Therefore, $T_{leaf}-T_{air}$ decreases with $g_a$, which in turn increases with wind speed [Fig. `r fig_leaf_T`b,@daudetWindSpeedLeaf1999a] and decreases with leaf size (Fig. `r fig_leaf_T`d);
decreases with $g_s$ (Fig. `r fig_leaf_T`e);
and decreases with vapor pressure deficit, or increases with RH (Fig. `r fig_leaf_T`c).

>>>>

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
Foliage distribution is structured in correspondence with the distribution of tree sizes within a forest, where leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs. `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`). 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In open canopy forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig. `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations followed by disturbance all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019]. 
In this review, we focus on growing season conditions (peak LAD) unless otherwise noted.

**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the profile shape modified by LAD, canopy height, canopy structure, and across species and forest types [Figs. `r fig_NEON_vertical`b-d, `r SIfig_NEON_all`, @koike_leaf_2001].** 
This gradient is more pronounced under dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), then under more open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs. `r fig_NEON_vertical`d, `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].

**Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018] to open forests and savannas [@curtis_intracanopy_2019; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs. `r fig_NEON_vertical`e, `r SIfig_NEON_all`). 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to 4.1 m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds averaged $\le$ 1.1 m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).

**Atmospheric concentrations of carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.**
Atmospheric CO~2~ concentrations tend to be higher near ground level at night, associated with plant and soil respiration, but the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999]. Elevated CO~2~ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are negligible for most of the day when photosynthesis is occurring, gradients in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.

Relative humidity, RH, tends to be higher in the understory and decreases with height across the vertical profile, although this trend is absent in open forests [Figs. `r fig_NEON_vertical`f, `r SIfig_NEON_all`, e.g., @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018].
Dense-canopy forests maintain higher daily maximum RH below-canopy than open forests and nearby open areas, and this effect is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a]. 
In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].

@@

**The buffering patterns described above vary geographically, seasonally, and across forest types [Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`, @frenneForestMicroclimatesClimate2021].**
The strength of buffering increases with canopy density -- for example, across gradients of elevation [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @juckerCanopyStructureTopography2018], moisture, and proximity to water, for example in riparian forests [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019].
Seasonally, canopy buffering capacity is greater in wet than dry seasons, presumably because leaf area is greater in wet seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a].

@@

Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections) with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020]. 
Leaves can be either cooler than air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ [megathermy, @doughtyAreTropicalForests2008].
While $T_{leaf}$ is rarely exactly equal to $T_{air}$ (poikilothermy), it is most commonly within a few degrees [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

@@

**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based via biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input from radiation minus heat lost to the environment [Fig. `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].**
The dominant component of a leaf's energy balance is net radiation ($R_n$, including shortwave and longwave). High radiation loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig. `r fig_leaf_T`a).
Generally, the dominant avenue of heat loss to the environment is latent heat flux ($\lambda E$), wherein the loss of energy required to convert water from a liquid to a vapour state during evaporation and transpiration cools the leaf. 
Evapotranspiration ($E$) is controlled primarily by stomatal and boundary layer conductances, $g_s$ and $g_a$, and by the water vapour concentration difference between the leaf surface and the bulk air.
Therefore, $T_{leaf}-T_{air}$ decreases with $g_a$, which in turn increases with wind speed [Fig. `r fig_leaf_T`b,@daudetWindSpeedLeaf1999a] and decreases with leaf size (Fig. `r fig_leaf_T`d);
decreases with $g_s$ (Fig. `r fig_leaf_T`e);
and decreases with VPD, or increases with RH (Fig. `r fig_leaf_T`c).
